---
title: "Estimate Standardized Incidence (Mortality) Ratio (SIR/SMR) using R"
author: "Kossi D. ABALO"
date: 
output: html_document
---


<!--- for tables outpout, I used knitr and kableExtra packages. For more documentation :
http://haozhu233.github.io/kableExtra/awesome_table_in_html.html
To install kableextra package use: 

install.packages("kableExtra")

# For dev version
# install.packages("devtools")
devtools::install_github("haozhu233/kableExtra")
 --->


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r message=FALSE, warning=FALSE}
rm(list=ls())
library(kableExtra)
library(magrittr)
library(dplyr)
library(tidyverse)
library(popEpi)
library(DiagrammeR)
library(biostat3)
library(Epi)
```

# Introduction


Standardized Incidence Ratio (SIR) or Standardized Mortality Ratio (SMR) are common measure of event occurrence in epidemiology. They are used to compare the occurrence of an event in a cohort to that observed in a given population called reference population.  
SIR or SMR help the investigators to have a global idea on the occurrence of the event of interest in the followed population (the cohort).
They are an indirect method of adjustment for a confounding factors (age, sex, calendar year etc.) that describes in numerical terms how the cohort average experience of the event during the follow-up compared with that of the reference population as a whole. A $SIR=2$ means that the occurrence of the event event in the cohort is 2 times higher than what is expected in the reference population. But caution, this does not mean any causality in regard of a give exposure in the cohort or a given risk factor for which the cohort is followed for.

Mathematically SIR (SMR) is the ratio of observed events and expected events: 

$$SIR = \frac{\sum d_j }{\sum n_j \lambda_j} = \frac{D}{E}$$

where $D$ is the observed events in the cohort population and $E$ is the expected number of events.

Observed events are the absolute number of events that occurred in the cohort during the follow up.

The expected number of events are derived by multiplying the cohort person-years with reference **population rate**. The **population rate** should be stratified or adjusted by confounding factors (age or age group, gender, calendar period etc.). The reference **population rate** in strata $i$ ($\lambda_i$) is defined as: 
$$\lambda_i = \frac{d_i}{n_i}$$ 
where $d_i$ is the total observed events and $n_i$ is the total observed person years (or the size of the population) in the $i$th strata of the reference population.

Univariate confidence intervals are based on exact values of Poisson distribution and the formula for p-value is $$ \chi^2 = \frac{ (|O - E| -0.5)^2 }{E}. $$ Modeled SIR is a Poisson regression model with log-link and cohorts person-years as a offset (we will come to that in the next sections).

Let's have a quick look to some examples of cohort and reference data:

```{r include=FALSE}

cohort <- as.data.frame(cbind(
  age_group = c("35-39", "40-44", "45-49", "50-54", "55-59", "60-64", "65-69", "70-74"), 
  Number_events = c(0, 1, 3, 5, 8, 8, 4, 1),
  Pers_years = c(480, 587, 680, 541, 479, 356, 157, 36)))

pop <- as.data.frame(cbind(
  age_group = c("35-39", "40-44", "45-49", "50-54", "55-59", "60-64", "65-69", "70-74"), 
  Number_events = c(81, 177, 336, 434, 419, 599, 907, 925),
  Pers_years = 3*c(27575, 28670, 26970, 21190, 13625, 17275, 17515, 14075)))

class(pop$Pers_years) <- "numeric"
class(pop$Number_events) <- "numeric" 
class(cohort$Pers_years) <- "numeric"
class(cohort$Number_events) <- "numeric" 

as.numeric(as.character(pop$Number_events))

pop$Rate <- pop$Number_events/pop$Pers_years

```

```{r}
pop$Rate
pop$Rate*100000
```



```{r Table1, echo=FALSE}
cohort %>%
  kbl(caption = "Table 1: Example of cohort data") %>%
  kable_classic(full_width = F, html_font = "Cambria")
```
In this first example @\ref(table:Table1): 

  + *age_group* is the age group (confounding factor) for adjustment. Here we have only one confounding factor but feel free to add any other factors (age, gender, socio-economic factors, location, years etc.) you think necessary for your analyses and on which yours SIR/SMR will be adjusted. 
  + The *Number_events* is the total number of events observed in the cohort by strata.
  + The *Pers_years* is the total person-time recorded in the cohort for each strata.


```{r Table2, echo=FALSE}
pop %>%
  kbl(caption = "Table 2: Example of population reference data") %>%
  kable_classic(full_width = F, html_font = "Cambria")
```

In this second example [Table 2](#Table2) : 

  + *age_group* is the age group (confounding factor) for adjustment in the reference population. 
  + The *Number_events* is the total number of events observed in the reference population $n_i$ for each strata of *age_group*.
  + The *Pers_years* is the total the size of the reference population $n_i$ for each strata of *age_group*. 
  + *Rate* is the rate of the event $\lambda_i$=$n_i$/$d_i$  in the reference population for each strata of *age_group*. 

```
Note: make sure the same variables with with the same labels and the same classes in both your cohort data and in your population or reference data.
```
From the [Table 2](#Table2), we can extract the Rate column to merge with the cohort data [Table 1](#Table1) and we can calculate the *Expected* number of events obtained as population reference *Rate* in the $i$th strata multiplied by the cohort $j$th strata *Pers_years*. This is expressed as : 

$Expected$ = $n_i$*$\lambda_j$

```{r include=FALSE}
cohort %<>% 
  inner_join(pop[ , c(1, 4)]) %>% 
  mutate(Expected = Rate*Pers_years)
```

```{r Table3, echo=FALSE}

list(cohort[ , c(1:3)], cohort[ , c(4, 5)]) %>%
  kbl(caption = "Table 3: Calculate the expected number of cases in the cohort data", valign = "t", vline = " ", table.attr ='style="color:red"' ) %>%
  kable_classic(full_width = F, html_font = "Cambria")


```

The final step is now to calculate the SIR value as sum of Number_events divided by the sum of number of Expected :

\begin{equation} 
\begin{split}
  SIR &=\frac{\sum d_j }{\sum n_j \lambda_j} \\
& = \frac{D}{E} \\
& =\frac{(0+1+3+5+8+8+4+1)}{(0.4699909+1.207987+2.823878+3.693472+4.910116+4.114694+2.710039+0.7886323)} \\
& = 1.44796
\end{split}
\end{equation} 

```{r include=FALSE}
cat(cohort$Expected, sep = "+")
```


```{r include=FALSE}
sum(cohort$Number_events)/sum(cohort$Expected)

(0+1+3+5+8+8+4+1)/(0.4699909+1.207987+2.823878+3.693472+4.910116+4.114694+2.710039+0.7886323)

```

<font color=blue> 
**Interpretation : **

In our example, $SIR = 1.45$ means that standardizing on the age group, incidence in our studied cohort is **1.45 times** higher than in the reference population.
</font>

# How to implement SIR analysis on R? 
In this post, I will presente you three different way of SIR analyses on R. 
 
First of all, let me remember you that raw data for a given study are never in the form we saw in the [Table 1](#Table1) and [Table 2](#Table2). Most the time, you ought to configure your data in way to obtain [Table 1](#Table1) and [Table 2](#Table2) as in our examples. I named this step of data manipulation or data configuration as a data management step. 

Overall I will presente you the following steps: 


```{r echo=FALSE}
DiagrammeR::grViz("digraph {
  graph [layout = dot, rankdir = TB]
  
  node [shape = rectangle]        
  rec1 [label = 'Step 1. Data Management']
  rec2 [label = 'Step 2. First Example']
  rec3 [label =  'Step 3. Second Example']
  rec4 [label = 'Step 4. Third Example']
  
  # edge definitions with the node IDs
        
  
  rec1 -> rec2 
  rec1 -> rec3
  rec1 -> rec4
  }", 
  height = 100)
```

## 1 : Data management

From this step up to the end, I will be using a simulated cohort data named *mySIR* and a simulated population reference data *my.pop.ref*.

Using these data allows us only to compute SMR analysis with the research question: 

**is the cancer incidence rate among females diagnosed with rectal cancer the same as that in the general population?**

The cohort data simulated *mySIR* is on cancer occurrence in patients undergoing medical diagnostic ionizing radiation exposure from between 2000-2015. 

These are variables from the *mySIR* database :

- sex: the gender of the patient (1 = Male, 2 = female)

- birth_date: the date of birth (date: dd-mm-yyyy)

- entry_date: the date of exposure to ionizing radiation (date: dd-mm-yyyy)

- exit_date: the date of exit from follow-up (death or censoring) (date: dd-mm-yyyy)

- status: the status of the person at exit; 0 exit (or censored) without cancer; 1 diagnosed with cancer

```{r include=FALSE}
load("~/OneDrive - Aix-Marseille Université/Année 1 Thèse/Thèse Kossi/Analyse SIR/Analyse_SIR_V3_Article_Cohorte_Profile/base_these_kossi.Rdata")

mySIR=base_sans_doublon[ , c("sexe", "ddn", "date_entree", "dd_exit_suivi", "statut")]
colnames(mySIR) <- c("sex", "birth_date", "entry_date", "exit_date", "status")
mySIR %<>% 
  mutate(sex=ifelse(sex=="Male", 1, 2))

```

```{r echo=TRUE}
glimpse(mySIR) 
head(mySIR)
```
Overview of *mySIR* dataset : 
```{r echo=FALSE}
DT::datatable(mySIR, filter = 'top', options = list(pageLength = 5, autoWidth = TRUE))
```

From these variables in the dataset, let's create the following variables: 

- age: the age at exit from the cohort (ex_date - bi_date).

- year: the calendar year at exit from the cohort (ex_date - dg_date).

- Follow_up : the duration of follow-up in years of each patient of the cohort (exit_date - entry_date).

- age_exit : age at exit from the follow-up in years (exit_date - birth_date).

```{r}

mySIR %<>% 
  mutate(year=format(exit_date, format="%Y"),# to extract only the year (YYYY) of exit from the follow-up
         age_exit= cal.yr(exit_date) - cal.yr(birth_date), # using the function cal.yr from the package "Epi" to compute the age at exit from  the follow-up in years.
        age = cut(age_exit, breaks = c(0, 1, 5, 10, 15, 20), # To create age groups, coded 1 for <1 year, 2 for 1 to 4 years, 3 for 5 to 9 years, 4 for 10 to 14 years and 5 for >= 15 years old. 
                  labels = c(1, 2, 3, 4, 5), right = FALSE), 
        Follow_up = (exit_date - entry_date)/365.25) # using the function cal.yr from the package "Epi" to compute the Follow-up in years.
```


The reference population data simulated *my.pop.ref* is on cancer incidence in the general population from 2000-2015, by age group and by gender. The followings could be found from the dataset :

- sex: the gender of the patient (1 = Male, 2 = female)

- year: calendar year (format "YYYY")

- age: patients age group at cancer diagnosis, coded 1 for <1 year, 2 for 1 to 4 years, 3 for 5 to 9 years, 4 for 10 to 14 years and 5 for >= 15 years old.

- Number_cases: the total number of cases ($n_i$) in each age, year and sex strata of the reference population data

- Population: the size of the population ($d_i$) in each age, year and sex strata of the reference population data

- Rate: the average population cancer incidence rate per person-year ($\lambda_i$=$n_i$/$d_i$), where $n_i$ is the number of deaths and $d_i$ is the person-years). $n_i$ and $d_i$ are not explicitly in the data set, we only the resultant rate $\lambda_i$.  



```{r include=FALSE}
#link <- "C:/Users/antoi/OneDrive - Aix-Marseille Université/Année 1 Thèse/Thèse Kossi/Analyse SIR/TestProc/Test/ANALYSES_SUR_LA_V8/BASES_ORIGINALES/Taux_de_reference_France_Final.xlsx"

#ref <- readxl::read_excel(link)

ref <- readxl::read_excel("C:/Users/antoi/Desktop/000_COCCINELLE/DONNEES REGISTRES/DONNEES RNCE/Taux d'incidence pour la France 2000 à 2015/TI_RNCE_20002015.xlsx",
                               sheet = "TI_RNCE")
ref %>% 
  mutate(haz = `TI (cas/million)`/1000000, 
               Population = `Nb cas`/ haz) %>% 
  rename(RNCE_GROUPE_ICCC = `groupe ICCC`, 
         age = `Classe age`,
         Number_cases = `Nb cas`,
         year = annee,
         sex = sexe) %>% 
  filter(RNCE_GROUPE_ICCC != "LCH") %>% 
  dplyr::select(sex, year, age, Number_cases, Population) -> mypopref


popu <- mypopref[which(!duplicated(mypopref[c("sex", "year", "age")])), 
                c("sex", "year", "age", "Population")]

mypopref %>% 
  group_by(sex, year, age) %>% 
  summarise(Tot_cas = sum(Number_cases)) -> cases


popu %>% 
  group_by(sex, year, age) %>% 
  summarise(Tot_Pers_year = sum(Population)) -> Pers.year



my.pop.ref <- inner_join(Pers.year, cases) %>% 
  mutate(Rate=Tot_cas/Tot_Pers_year)

rm(mypopref, popu, cases)

colnames(my.pop.ref) <- c("sex", "year", "age", "Population", "Number_cases", "Rate")

```


```{r}
glimpse(my.pop.ref)
head(my.pop.ref)
```

Overview of *my.pop.ref* dataset : 
```{r echo=FALSE}
DT::datatable(my.pop.ref, filter = 'top', options = list(pageLength = 5, autoWidth = TRUE))
```


### Summarize the cohort data
<!---
```{r}
mySIR %>% 
  group_by(sex, year, age) %>% 
  summarise(Pers.years = sum(Follow_up)) -> mySIR.Pers.years # To calculate Person-time for each sex, year and age group strata


mySIR %>% 
  group_by(sex, year, age) %>% 
  summarise(Observed = sum(status)) -> mySIR.total.events #  To calculate total number of observed events recorded in each sex, year and age group strata
mySIR.summarized <- inner_join(mySIR.Pers.years, mySIR.total.events)

```

```{r include=FALSE}
sum(mySIR.total.events$Observed)
rm(mySIR.Pers.years, mySIR.total.events)
```

Summarizing our cohort data help us to have a table like the [Table 1](Table1).

```{r Table4, echo=FALSE}
mySIR.summarized[1:10, ] %>%
  kbl(caption = "Table 4: Summarized cohort data") %>%
  kable_classic(full_width = F, html_font = "Cambria")
```

The *Observed* column presents the total number of observed events recorded and the *Pers.years* column presents the total number of person-time recorded in each sex, year and age group strata.
--->


We can also obtain the summary by using the function *lexpand* of the package *popEpi* which allows quickly summary of cohort data. 

```{r}
cohort_lex <- dplyr::select(mySIR, -c("age", "age_exit"))
lex.sir <- lexpand(cohort_lex,
                   birth = birth_date,
                   entry = entry_date,
                   exit = exit_date,
                   status = status==1,
                   breaks = list(fot=0:20, # cut the follow-up time in periods of one year
                                 age=c(0, 1, 5, 10, 15, 20), # cut the age at exit in subgroups, coded 1 for <1 year, 2 for 1 to 4 years, 3 for 5 to 9 years, 4 for 10 to 14 years and 5 for >= 15 years old. 
                                 per=2000:2020), # To cut the calendar year in periods of one year. You can choose also to cut in subgroups of 5 or 10 years 
                   aggre = list(sex,
                               age,
                               year=per)# rename the per variable in year as in the population data
                   )
glimpse(lex.sir)
```

The object *lex.sir* is an aggregated cohort data. The original cohort data *mySIR* is splited by the function *lexpand* into to subintervals of time over calendar time (year), age, and follow-up time (fot) with given time breaks using *splitMulti*.
lex.sir

In the *lex.sir* object, the  *from0to1* column represents the total number of observed events recorded and the *pyrs* column represents the total number of person-time recorded in each sex, year and age group strata.


## 2: First Example of SIR computation

Since we computed our cohort data and population reference data similar to tables 1 and 2, we can now easily calculate the SIR in our cohort according to the first equation.

Remember that both the population reference data *my.pop.ref* and the cohort data *lex.sir* should contain the same column names with the same classes and the same codifications for at least the variables of adjustment (sex, age, year).


- Calculate the expected value

<!---
```{r eval=FALSE, include=FALSE}
mySIR.summarized$year = as.numeric(as.character(mySIR.summarized$year))
mySIR.summarized$age = as.numeric(as.character(mySIR.summarized$age))
#data.sir <- inner_join(mySIR.summarized, my.pop.ref[ , c("sex", "year", "age", "Rate")])
data.sir <- merge(mySIR.summarized, my.pop.ref[ , c("sex", "year", "age", "Rate")], by=c("sex", "year", "age"), all.x = TRUE) # To merge the cohort data with the population rate

data.sir$Expected = data.sir$Pers.years*data.sir$Rate
```
--->

```{r include=FALSE}
lex.sir%<>% 
  mutate(age=ifelse(age==0, 1, 
                     ifelse(age==1, 2,
                            ifelse(age==5, 3, 
                                   ifelse(age==10, 4, 
                                          ifelse(age==15, 5, NA)))))) 
```

To calculate the SIRs, we will now merge by "sex", "year" and "age", the population reference data *my.pop.ref* and the cohort data *lex.sir* and then create a new column "Expected" as the Rate that multiply person-time.

```{r echo=TRUE}
lex.sir.final <- merge(lex.sir, my.pop.ref[ , c("sex", "year", "age", "Rate")],
                  by=c("sex", "year", "age"), all.x=TRUE)
lex.sir.final$Expected <- lex.sir.final$Rate*lex.sir.final$pyrs

```

Now since we have the "Observed" and "Expected" columns ready in our dataset, we can calculate the overall SIR as the sum of "Observed" divided by the sum of "Expected".

- Overall SIR
```{r echo=TRUE}
sum(lex.sir.final$from0to1, na.rm = TRUE)/sum(lex.sir.final$Expected, na.rm = TRUE) # Exclude NA values
```

We can also calculate SIRs in different ways and for different variables as follow :

- SIRs by gender 
```{r echo=TRUE}
lex.sir.final %>% 
  group_by(sex) %>% 
  summarise(O=sum(from0to1),
            E=sum(Expected, na.rm = TRUE),
            sir_sex=sum(from0to1)/sum(Expected, na.rm = TRUE))

```

- SIRs by age group 
```{r echo=TRUE}
lex.sir.final %>% 
  group_by(age) %>% 
  summarise(O=sum(from0to1),
            E=sum(Expected, na.rm = TRUE),
            sir_age=sum(from0to1)/sum(Expected, na.rm = TRUE))

```

- SIRs by calendar year 
```{r echo=TRUE}

lex.sir.final %>% 
  group_by(year) %>% 
  summarise(O=sum(from0to1),
            E=sum(Expected, na.rm = TRUE),
            sir_year=sum(from0to1)/sum(Expected, na.rm = TRUE))
```



## 3: Second Example of SIR computation

Since observed cases in the cohort are a typical case of Poisson distribution, Poisson model can be used to estimate the SIRs as follow:

- Overall SIR

```{r}
fit=glm(from0to1~offset(log(Expected)), data=lex.sir.final, family = "poisson")
eform(fit)
```


- SIRs by gender

```{r echo=TRUE}
fit=glm(from0to1[sex==1]~offset(log(Expected[sex==1])), data=lex.sir.final, family = "poisson") # for males
eform(fit)

fit=glm(from0to1[sex==2]~offset(log(Expected[sex==2])), data=lex.sir.final, family = "poisson") # females
eform(fit)
```

<font color=green > 
**Using Poisson regression, the SIRs are little bit reduced (underestimated). But I would not be able to explain why do we have such difference.**
</font> 


We can also use Poisson test to compute our SIRs

- By gender
```{r echo=TRUE}
by(lex.sir.final, lex.sir.final$sex, function(data) poisson.test(sum(data$from0to1), sum(data$Expected, na.rm = TRUE))) 

```

- By age group
```{r echo=TRUE}
by(lex.sir.final, lex.sir.final$age, function(data) poisson.test(sum(data$from0to1), sum(data$Expected, na.rm = TRUE)))

```

- By year

```
by(lex.sir.final, lex.sir.final$year, function(data) poisson.test(sum(data$from0to1), sum(data$Expected, na.rm = TRUE)))
```


## 4: Third example of SIR computation

SIR using the function *sir* of the *popEPi* package:

- by sex
```{r echo=TRUE}
sr <- popEpi::sir(lex.sir.final, coh.obs = 'from0to1',
          coh.pyrs = 'pyrs',
          ref.data = my.pop.ref[ , c("sex", "year", "age", "Rate")],
          ref.rate = Rate,
          print = c("sex"),
          adjust = c("age", "sex", "year"),
          test.type = "homogeneity",
          conf.type = "wald",
          conf.level = 0.95, EAR = F)


sr
```

- by age group

```{r echo=TRUE}
sr <- popEpi::sir(lex.sir.final, coh.obs = 'from0to1',
          coh.pyrs = 'pyrs',
          ref.data = my.pop.ref[ , c("sex", "year", "age", "Rate")],
          ref.rate = Rate,
          print = c("age"),
          adjust = c("age", "sex", "year"),
          test.type = "homogeneity",
          conf.type = "wald",
          conf.level = 0.95, EAR = F)


sr
```








